---
title: "Diagonal Approximation of LD matrix"
author: "Nicholas Knoblauch"
date: 2017-08-04
output: workflowr::wflow_html
---





#Introduction


The LD matrix $R$ is banded. It's possible to approximate a banded matrix as block-diagonal. In theory, as the block size gets smaller, the approximation should get worse. For this analysis, I've run 500 simulations with $PVE=0.1$ and $c=0.01$.  Using 18 block sizes ranging from $p$ (12758), to 10, I approximate the full LD matrix as block diagonal, take the EVD of each block, and instead of maximizing a single multivariate normal (log) density function that has a  $p\times p$ covariance matrix,the optimization is now maximizing the sum of $K$ multivariate densities, where $K$ is $\lceil  \frac{p}{\text{chunksize}} \rceil$.

The plot below shows the distribution of RMSE relative to the RMSE calculated for $\text{chunksize}=p$. The y axis represents chunks of decreasing size (chunksize is plotted as $\frac{\text{chunksize}}{p}$).  As you can see, chunking doesn't seem to matter that much.




```{r,echo=F,message=F,warning=F}
library(LDshrink)



library(RcppEigenH5)
small_genof <- "/home/nwknoblauch/Dropbox/eqtl_estimation/data/RSS_examples/genotype2.mat"
small_evdf <- "/home/nwknoblauch/Dropbox/eqtl_estimation/data/small_sim_evd.h5"
small_simf <- "../data/polygenic_sim_genotype2/simulation2.RDS"
large_genof <- "/home/nwknoblauch/Downloads/genotype.mat"

large_evdf <- "/home/nwknoblauch/Dropbox/eqtl_estimation/data/large_sim_evd.h5"
large_evdf <- "/home/nwknoblauch/Dropbox/eqtl_estimation/data/large_diag_approx_evd.h5"
#largest_simf <- "../data/polygenic_sim_genotype_larger/simulation.RDS"
large_simf <- "../data/polygenic_sim_genotype/simulation2.RDS"
rep_simf <- "../data/rep_polygenic/simulation.RDS"


library(RSSp)
SNP  <- scale(t(read_2d_mat_h5(large_genof,"/","C")),center=T,scale=F)
outdir <- "../data/rep_polygenic/"
nreps <- 500
pve <- 0.1
bias <- 0.01
nasimf <- gen_sim(SNP,pve,bias,nreps =nreps,outdir = outdir,collapse_all = T)


asimf <- "../data/rep_polygenic/simulation.RDS"

evdf <- large_evdf
simf <- large_simf
genof <- large_genof
p <- get_colnum_h5(large_genof,"/","R")
n <- get_colnum_h5(large_genof,"/","C")

```

## Direct Simulation

By simulating data using $R$ directly, we can better see the effect of chunking on estimating parameters
```{r,echo=F,eval=F}
dir_simf <- "../data/polygenic_sim_genotype/simulation4.RDS"
chunksizes <- unique(c(as.integer(seq.int(p,100,length.out = 10)),
                       as.integer(seq.int(1506,100,length.out = 10))))



dres_dfl <- list()
#res_dfl[[1]] <- test_res_df
for(i in 1:length(chunksizes)){
  cat(i,"\n")
  dres_dfl[[i]] <- estimate_RSSp_files(evdf,dir_simf,genof,chunksizes[i])
}
dares_df <- bind_rows(dres_dfl)
saveRDS(ares_df,"../output/rep_chunk_sim_direct.RDS")
```




```{r,echo=F,message=F,warning=F}
library(tidyverse)
library(ggjoy)



dares_df <- readRDS("../output/rep_chunk_sim_direct.RDS") %>% mutate(rmse=abs(pve-tpve))
dtparamdf <- distinct(dares_df,tpve,tbias,tsigu)
# ggplot(bad_resdf,aes(x=chunksize,y=rmse,col=fgeneid))+geom_point()+geom_smooth(method="lm",se=F)
mrmse <- group_by(dares_df,tpve,tbias,chunksize) %>% summarise(mean_pve_rmse=mean(rmse),sd_rmse=sd(rmse)) %>% ungroup()

ggplot(mrmse,aes(x=chunksize,y=mean_pve_rmse))+geom_point()+geom_smooth()
dnres_df <- group_by(dares_df,fgeneid) %>% filter(chunksize==max(chunksize)) %>% mutate(denom_rmse=rmse) %>% ungroup() %>% select(fgeneid,denom_rmse) %>% inner_join(dares_df) %>% group_by(fgeneid) %>% mutate(rel_rmse=rmse/denom_rmse,rel_dim=chunksize/p)

bad_resdf <- group_by(dares_df,fgeneid) %>% summarise(max_rmse=max(rmse)) %>% filter(max_rmse>0.75) %>% inner_join(dares_df)

filter(dnres_df,rel_dim<1) %>% ggplot(aes(x=rel_rmse,y=factor(rel_dim)))+geom_joy()+ggtitle("Relative RMSE vs chunksize","Foldchange over no chunking")+scale_x_log10()+xlab("Foldchange RMSE")+ylab("Relative Chunksize")

```

Zooming in on the region around 1:

```{r}
filter(dnres_df,rel_dim<1) %>% ggplot(aes(x=rel_rmse,y=factor(rel_dim)))+geom_joy(rel_min_height=0.01)+ggtitle("Relative RMSE vs chunksize","Foldchange over no chunking")+scale_x_log10(limits=c(0.1,10))+xlab("Foldchange RMSE")+ylab("Relative Chunksize")
```




```{r}
# denom_rmse <- filter(ares_df,chunksize==max(chunksize)) %>% summarise(mean_denom_rmse=mean(rmse))
# sum_df <-  group_by(ares_df,chunksize) %>% summarise(mean_rel_rmse=mean(rmse)/denom_rmse$mean_denom_rmse)
ggplot(dnres_df,aes(x=(chunksize^2)/p^2,y=rel_rmse))+geom_point()+geom_smooth()+scale_x_log10()+scale_y_log10()
```





