---
title: "Marginalized Likelihood Curves"
author: "Nicholas Knoblauch"
date: 2018-04-12
output: workflowr::wflow_html
---

# Likelihood curves for RSSp

```{r,echo=F,message=F,warnings=F}
library(EigenH5)
library(tidyverse)
library(RSSp)
quhf <- "/home/nwknoblauch/Desktop/scratch/polyg_scratch/RSSp_genome_gwas_quh/chr1-22_genred2_genred2_T_omni_AllPVE_T.h5"
estf <- "/home/nwknoblauch/Dropbox/PolygenicRSS/output/RSSp_snakemake/sim_chr1-22_genred2_genred2_T_omni_gwas_RSSp_res_AllPVE_1_T_F.txt.gz"
res_df <- read_delim(estf,delim="\t") %>% select(optim_pve=pve,tpve,fgeneid) %>% mutate(fgeneid=as.character(fgeneid))
tparam_df <- read_df_h5(quhf,"SimulationInfo")
tparam_df <- group_by(tparam_df,tpve) %>% mutate(pve_replicate=1:n()) %>% ungroup()

#p_n <- unique(tparam_df$p/tparam_df$n)

#quhm <- read_matrix_h5(quhf,"/","quh")
#D <- read_vector_h5(quhf,"/","D")
grid_df <- readRDS("out_res.RDS")
#grid_df <- array_branch(quhm,2) %>% map2_df(tparam_df$fgeneid,~RSSp_estimate_grid(v_hat = .x,eigenvalues = D,p_n = p_n,grid_points = 100)%>% mutate(fgeneid=.y)) %>% inner_join(tparam_df)

#saveRDS(grid_df,"out_res.RDS")

#lnZ_good <- RSSp_estimate_grid(v_hat = quhm[,2],eigenvalues = D,p_n = p_n,grid_points = 50) %>% mutate(fgeneid="21") %>% inner_join(tparam_df)  
```


## Likelihood curves

For each simulation (`pve_replicate`), I calculated the log-likelihood for 100 values of $PVE$ evenly spaced between 0 and 1.  To aid visualization, I subtracted the mean and divided by the standard deviation for each simulation to obtain the **Normalized Log-Likelihood**.  The vertical colored bar corresponds to the MLE obtained by `optim`, and the black vertical bar corresponds to the true $PVE$ value. What you'll notice is that 1) the shape of the likelihood curves can vary greatly and 2) `optim` consistently finds the "bottom" of the curve, and (most importantly) 3) the "bottom" of the curve is sometimes quite far from the true PVE value. 


```{r}
filter(grid_df,fgeneid %in% c("16","21")) %>% mutate(PVE_replicate=factor(pve_replicate)) %>% rename(True_PVE=tpve) %>% ggplot(aes(x=pve,y=rel_lnZ,col=PVE_replicate))+
  geom_point()+
  geom_vline(aes(xintercept = True_PVE),linetype=2)+xlab(bquote(PVE))+ylab("Normalized Log Likelihood")+ggtitle("'True' PVE is sometimes far from the minimum","True PVE along black dotted line")+
  facet_wrap(~True_PVE,scales = "free_y",labeller = "label_both")+
  geom_vline(aes(xintercept=optim_pve,col=PVE_replicate))
```


And here's the plot for several more simulations

```{r}
grid_df <- group_by(grid_df,fgeneid) %>% mutate(rel_lnZ=(lnZ-mean(lnZ))/sd(lnZ)) %>% ungroup() %>% inner_join(res_df)
mutate(grid_df,PVE_replicate=factor(pve_replicate)) %>% ggplot(aes(x=pve,y=rel_lnZ,col=PVE_replicate))+
  geom_point()+
  geom_vline(aes(xintercept = tpve),linetype=2)+xlab(bquote(PVE))+ylab("Normalized Log Likelihood")+ggtitle("Normalized Log likelihood along with MLE","True PVE along black dotted line")+
  facet_wrap(~tpve,scales = "free_y")+
  geom_vline(aes(xintercept=optim_pve,col=PVE_replicate))
```


