---
title: "Out Of Sample LD"
author: "Nicholas Knoblauch"
date: 2018-04-12
output: workflowr::wflow_html
---

# Out of Sample LD

For these simulations I'm using the wtccc `cd` genotypes for GWAS simulation and the `bd` genotypes as a reference panel for the purposes of calculating LD.  

##  Simulation details




```{r libs,echo=F,message=F,warnings=F}
library(EigenH5)
library(tidyverse)
library(plotly)
```



```{r load_d,echo=F,message = F,warnings = F}
res_d_rssp <- dir("~/Dropbox/PolygenicRSS/output/RSSp_snakemake",full.names = T,pattern="SNP")
res_d_ldsc <- dir("~/Dropbox/PolygenicRSS/output/RSSp_snakemake/ldsc_res/",pattern="cd",full.names = T)

res_df_rssp <- map_df(res_d_rssp,~quietly(read_delim)(.x,delim = "\t")$result %>% mutate(chrom=as.character(chrom),method="RSSp",useLDetect=as.character(useLDetect))) %>% filter(!is.na(simulation),!is.na(useLDshrink)) %>% filter(scenario == "SmallvPVE", dataset == "cd", panel_dataset == "bd",chrom=="1-22",simulation=="gwas",!is.na(chrom),!is.na(AF)) %>% select(-ends_with("bound"),-sigu,-lnZ,-bias,-tbias,-optim,-useGradient,-log_params,-message,-convergence)
res_df_ldsc <- map_df(res_d_ldsc,~quietly(read_delim)(.x,delim = "\t")$result %>% mutate(Intercept=as.character(Intercept),chrom=as.character(chrom),method=ifelse(useIntercept,"LDSC_Intercept","LDSC_noIntercept"),useLDetect=as.character(useLDetect))) %>% filter(scenario == "SmallvPVE", dataset == "cd", panel_dataset == "bd",chrom=="1-22",simulation=="gwas",!is.na(chrom),!is.na(AF))%>% select(-Lambda_GC,-Mean_Chi_2,-Total_time_elapsed,-Intercept)
res_df <- bind_rows(res_df_rssp,res_df_ldsc)

sum_df <- res_df %>%  group_by_at(vars(-pve,-fgeneid)) %>% summarise(mean_pve=mean(pve),sd_pve=sd(pve),error_pve=mean((pve-tpve)^2),error_sd=sd((pve-tpve)/tpve),rel_error_pve=mean(((pve-tpve)/tpve)^2)) %>% ungroup() %>% mutate(error_rmse=sqrt(error_pve))
```




```{r rmse_oos,echo=F,message = F,warnings = F}
p <- filter(sum_df,SNPCT==0,
            method!="LDSC_Intercept",((useLDshrink & method=="RSSp") | (!useLDshrink & method !="RSSp")))  %>% mutate(Method=gsub("(_.+)","",method)) %>% 
  ggplot(aes(x = tpve, y = error_rmse,fill=Method,col=Method)) + geom_errorbar(aes(ymin=error_rmse,ymax=error_rmse+error_sd/2),position=position_dodge())+ 
  geom_col(position=position_dodge()) +facet_grid(AF~N)+
  ggtitle("Root Mean Squared Error of Out-of-Sample-LD-based PVE Estimates","10 replicates per PVE value") +
  xlab("PVE") + 
  ylab("RMSE(PVE)")
p
```



```{r ldshrink_pve,echo=F,message = F,warnings = F}
p <- filter(sum_df,SNPCT==0,N==0,AF!=0.2) %>% ggplot(aes(x = tpve, y = mean_pve,col=method)) +
   geom_errorbar(aes(ymin=mean_pve-sd_pve/2,ymax=mean_pve+sd_pve/2))+
  geom_abline(slope = 1, intercept = 0) +geom_point()+
  facet_grid( AF~useLDshrink,labeller = "label_both") +
  ggtitle("RSSp vs LDSc across Allele Frequencies (with and without LDshrink)") +
  xlab("PVE") + 
  ylab("PVE Estimate")+scale_y_sqrt()+scale_x_sqrt()
print(p)
# ggplotly(p)
```


```{r ldshrink_improves,echo=F,message = F,warnings = F}
px <- filter(sum_df,SNPCT==0,N==0,AF!=0.2) %>% ggplot(aes(x = method, y = error_pve,fill=method)) +
  geom_boxplot() + geom_point()+
  facet_grid( AF~useLDshrink,labeller = "label_both",scales="free_y") +
  ggtitle("RSSp performance is vastly improved by LDshrink") +
  xlab("Method") + 
  ylab("PVE squared error")+
  scale_y_sqrt()
print(px)
#ggplotly(px)
```

### How much Better is `RSSp` than `LDSC`?





## Varying SNP count


```{r var_snp_pve,echo=F,message = F,warnings = F}
npx <- filter(sum_df,SNPCT!=0,N==0,AF!=0.2,useLDshrink) %>% ggplot(aes(x = tpve, y = mean_pve,col=method)) +
   geom_errorbar(aes(ymin=mean_pve-sd_pve/2,ymax=mean_pve+sd_pve/2))+
  geom_abline(slope = 1, intercept = 0) +geom_point()+
  facet_grid( AF~p,labeller = "label_both") +
  ggtitle("RSSp is robust across AF and SNP count") +
  xlab("PVE") + 
  ylab("PVE Estimate")
ggplotly(npx)
```


```{r var_snp_rmse,echo=F,message = F,warnings = F}
fpx <- filter(sum_df,SNPCT!=0,N==0,AF!=0.2,useLDshrink)  %>% ggplot(aes(x = method, y = error_pve,fill=method)) +
  geom_boxplot() + 
  facet_grid( AF~p,labeller = "label_both",scales="free_y") +
  ggtitle("RSSp outperforms LDSC across SNP count and Allele Frequency") +
  xlab("Method") + 
  ylab("PVE squared error")+
  scale_y_sqrt()
ggplotly(fpx)
```


## Varying Sample size

I also varied sample size.  In the `N=2000` scenario, both GWAS and reference panel genotypes were downsampled to `2000` individuals.  In the `N=4686` scenario, `N` reflects the size of the GWAS dataset, the reference panel dataset is `4806`.

```{r var_n_pve,echo=F,message = F,warnings = F}

np <- filter(sum_df,SNPCT==0,AF!=0.2,useLDshrink) %>% ggplot(aes(x = tpve, y = mean_pve,col=method)) +
   geom_errorbar(aes(ymin=mean_pve-sd_pve/2,ymax=mean_pve+sd_pve/2))+
  geom_abline(slope = 1, intercept = 0) +geom_point()+
  facet_grid( AF~n,labeller = "label_both") +
  ggtitle("RSSp is robust to changes in sample size") +
  xlab("PVE") + 
  ylab("PVE Estimate")
ggplotly(np)
```


```{r var_n_rmse,echo=F,message = F,warnings = F}
nnp <- filter(sum_df,SNPCT==0,AF!=0.2,useLDshrink)  %>% ggplot(aes(x = method, y = error_pve,fill=method)) +
  geom_boxplot() + 
  facet_grid( AF~n,labeller = "label_both",scales="free_y") +
  ggtitle("RSSp outperforms LDSC across sample size and Allele Frequency") +
  xlab("Method") + 
  ylab("PVE squared error")+scale_y_sqrt()
ggplotly(nnp)
```
