---
title: "Prediction with RSSp"
author: "Nicholas Knoblauch"
date: 2017-11-14
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Prediction with RSSp

Simulation was performed using ~700k SNPs across the genome. LD was computed block-wise, with blocks identified by LDetect.  Prediction is on held-out samples (from the same LD reference panel, but that were not used in calculation of LD).


```{r}
library(rhdf5)
library(purrr)
library(RcppEigenH5)
library(tidyverse)

est_f <- "/home/nwknoblauch/Dropbox/PolygenicRSS/output/RSSp_snakemake/sim_scz2_gwas_RSSp_res_ALL.txt.gz"
est_df <- read_delim(est_f,delim="\t") %>% mutate(fgeneid=as.character(fgeneid),
                                                  rel_bias=round(tbias/tpve,3),
                                                  true_pve=round(tpve,3)) %>% arrange(as.integer(fgeneid)) %>% 
  filter(log_params==F,
         useGradient==T,
         method=="Confound") %>%
  select(-log_params,-useGradient,-method)

oracle_f <- "/home/nwknoblauch/Dropbox/PolygenicRSS/output/RSSp_snakemake/sim_scz2_gwas_RSSp_oracle_res_ALL.txt.gz"
est_df <- read_delim(oracle_f,delim="\t") %>% mutate(fgeneid=as.character(fgeneid)) %>% select(fgeneid,oracle_lnZ) %>% inner_join(est_df)
tbetaf <- "/home/nwknoblauch/Desktop/scratch/polyg_scratch/RSSp_sim_gwas_beta/scz2_beta.h5"
ymat_estf <- "/home/nwknoblauch/Desktop/scratch/polyg_scratch/RSSp_sim_gwas_all_beta_est/cv_scz2_1.0_0.00031622776601683794_sim_beta.h5"
xmat_test <- "/home/nwknoblauch/Desktop/scratch/polyg_scratch/h5/scz2_seq_test_hapmap_geno.h5"
train_yf <- "/home/nwknoblauch/Desktop/scratch/polyg_scratch/RSSp_sim_gwas_pheno/scz2_trait.h5"
test_yf <- "/home/nwknoblauch/Desktop/scratch/polyg_scratch/RSSp_sim_gwas_pheno/scz2_trait_test.h5"

tparam_df <- est_df
beta_est <- read_2d_mat_h5(ymat_estf,"/","beta_est")
est_y <- read_2d_mat_h5(ymat_estf,"/","y_est")
oracle_y <- read_2d_mat_h5(ymat_estf,"/","y_oracle")

beta_oracle <- read_2d_mat_h5(ymat_estf,"/","beta_oracle")
beta_true <- t(read_2d_mat_h5(tbetaf,"/","beta"))
# test_X <- scale(t(read_2d_mat_h5(xmat_test,"/","dosage")),center=T,scale=F)
# se_files <- dir("/home/nwknoblauch/Desktop/scratch/polyg_scratch/RSSp_genome_gwas_uh_chunk",full.names = T,pattern="scz2")
# LDchunk <- as.integer(gsub(".+scz2_([0-9]+)_sim.h5","\\1",se_files))
# se_df <- data_frame(se_file=se_files,LDchunk=LDchunk) %>% arrange(LDchunk)
# se_files <- se_files
# all_se <- concat_mat_chunks(h5files = se_df$se_file,
#                             groupnames=as.character(se_df$LDchunk),datanames = rep("se",nrow(se_df)))



t_S <- read_vec(tbetaf,"S")
true_y <- read_2d_mat_h5(test_yf,"trait","ymat")


pheno_errors <- colMeans((est_y-true_y)^2)
true_beta_sigu <- apply(beta_true/t_S,2,sd)
oracle_beta_sigu <- apply(beta_oracle/t_S,2,sd)
est_beta_sigu <- apply(beta_est/t_S,2,sd)


adf <- array(c(true_y,est_y,oracle_y),dim = c(dim(true_y),3),dimnames=list(sample=as.character(1:nrow(true_y)),
                                                                           fgeneid=as.character(1:ncol(true_y)),
                                                                           method=as.character(c("true","est","oracle"))))

tblc <- as.tbl_cube(adf)
y_df <- as_data_frame(tblc) %>% mutate(fgeneid=as.character(fgeneid),method=paste0("y_",method)) %>% spread(key = method,value = adf) %>% inner_join(tparam_df)

# plot(pheno_errors,tparam_df$tpve)
errors <- colMeans((beta_est-beta_true)^2)
# errors <- colMeans((beta_est-beta_true)^2)

cor_beta_est_true <- map2_dbl(array_branch(beta_est,2),array_branch(beta_true,2),cor)
cor_beta_oracle_est <- map2_dbl(array_branch(beta_est,2),array_branch(beta_oracle,2),cor)
cor_beta_oracle_true <- map2_dbl(array_branch(beta_true,2),array_branch(beta_oracle,2),cor)

cor_y_est_true <- map2_dbl(array_branch(est_y,2),array_branch(true_y,2),cor)
cor_y_oracle_est <- map2_dbl(array_branch(oracle_y,2),array_branch(est_y,2),cor)
cor_y_oracle_true <- map2_dbl(array_branch(oracle_y,2),array_branch(true_y,2),cor)

result_df <- data_frame(beta_errors=errors,
                    pheno_errors=pheno_errors,
                    cor_beta_est_true=cor_beta_est_true,
                    cor_beta_oracle_est=cor_beta_oracle_est,
                    cor_beta_oracle_true=cor_beta_oracle_true,
                    cor_y_oracle_est=cor_y_oracle_est,
                    cor_y_oracle_true=cor_y_oracle_true,
                    cor_y_est_true=cor_y_est_true,
                    true_sigu=round(true_beta_sigu,3),
                    oracle_sigu=oracle_beta_sigu,
                    est_sigu=est_beta_sigu) %>% mutate(fgeneid=as.character(1:n()))

tparam_df <- inner_join(tparam_df,result_df)

```
## Prediction results using estimated $PVE$ and $c$ (confounding).

Overall, 

```{r y_true_y_est,echo=F}
ggplot(tparam_df,aes(x=factor(true_pve),y=cor_y_est_true))+geom_boxplot()+facet_grid(.~rel_bias,labeller = partial(label_both,sep=":\n"))+xlab(bquote(PVE))+ylab(bquote(rho(y[true]~','~tilde(x)~bar(beta)[post])))+ggtitle(bquote(Correlation~between~tilde(x)~bar(beta)[posterior]~and~y[true]))
```


```{r}
ggplot(y_df,aes(x=y_true,y_est))+geom_point()+facet_grid(true_pve~rel_bias,labeller=partial(label_both,sep=":\n"))+geom_smooth(method="lm")+xlab(bquote(y[true]))+ylab(bquote(y[posterior]))+ggtitle(bquote(Correlation~between~tilde(x)~beta[posterior]~and~y[true]))
```

## Prediction results using *true* $PVE$ and $c$ (confounding).


```{r}
ggplot(tparam_df,aes(x=factor(true_pve),y=cor_y_oracle_true))+geom_boxplot()+facet_grid(.~rel_bias,labeller = partial(label_both,sep=":\n"))+xlab(bquote(PVE))+ylab(bquote(rho(y[true]~','~y[oracle])))+ggtitle(bquote(Correlation~between~tilde(x)~bar(beta)[oracle]~and~y[true]))
```
```{r}
ggplot(y_df,aes(x=y_true,y_oracle))+geom_point()+facet_grid(true_pve~rel_bias,labeller=partial(label_both,sep=":\n"))+geom_smooth(method="lm")+xlab(bquote(y[true]))+ylab(bquote(y[oracle]))
```


```{r}
ggplot(tparam_df,aes(x=factor(true_pve),y=cor_y_oracle_est))+geom_boxplot()+facet_grid(.~rel_bias,labeller = partial(label_both,sep=":\n"))+xlab(bquote(PVE))+ylab(bquote(rho(y[oracle]~','~y[posterior])))+ggtitle(bquote(Correlation~between~tilde(x)~bar(beta)[oracle]~and~tilde(x)~bar(beta)[posterior]))
```


# Miscellaneous observations


1. The variance of the posterior mean of $\textbf{u}$ (conditioned on the *MLE* of $\sigma^2_u$ ) is much lower than the MLE of $\sigma_u^2$.  

```{r}
ggplot(tparam_df,aes(x=sigu,y=est_sigu))+geom_point()+xlab(bquote(sigma[u][MLE]))+ylab(bquote(SD(bar(u)[posterior])))+geom_abline(slope=1,intercept=0)+ggtitle("E[Var(u)]!=Var(E[u])","But they're highly correlated")+scale_x_log10()+scale_y_log10()

```


2. The variance of the posterior mean of $\textbf{u}$ (conditioned on the *true value* of $\hat{\sigma^2_u}$ ) is much lower than the *true value* of $\sigma_u^2$.  


```{r}
ggplot(tparam_df,aes(x=tsigu,y=oracle_sigu))+
  geom_point()+
  xlab(bquote(sigma[u][True]))+
  ylab(bquote(SD(u[Oracle])))+
  geom_abline(slope=1,intercept=0)+scale_x_log10(limits=c(1e-5,2e-2))+scale_y_log10(limits=c(1e-5,1e-2))
```


```{r}
ggplot(tparam_df,aes(x=factor(true_pve),y=cor_beta_est_true))+geom_boxplot()+facet_grid(.~rel_bias,labeller = partial(label_both,sep=":\n"))+xlab(bquote(PVE))+ylab(bquote(rho(bar(beta)[posterior]~','~beta[true])))+ggtitle(bquote(Correlation~between~bar(beta)[posterior]~and~beta[true]))
```



```{r}
ggplot(tparam_df,aes(x=factor(true_pve),y=cor_beta_oracle_true))+geom_boxplot()+facet_grid(.~rel_bias,labeller = partial(label_both,sep=":\n"))+xlab(bquote(PVE))+ylab(bquote(rho(bar(beta)[oracle]~','~beta[true])))+ggtitle(bquote(Correlation~between~bar(beta)[oracle]~and~beta[true]))
```

```{r}
ggplot(est_df,aes(x=tsigu,y=sigu))+geom_point()+geom_abline(slope=1,intercept=0)+xlab(bquote(sigma[u]))+ylab(bquote(hat(sigma)[u]))+facet_wrap(~rel_bias,labeller = partial(label_both,sep=":\n"))+xlim(0,0.026)+ylim(0,0.026)
```



```{r}
ggplot(est_df,aes(x=tpve,y=pve))+geom_point()+geom_abline(slope=1,intercept=0)+xlab(bquote(PVE))+ylab(bquote(hat(PVE)))+facet_wrap(~rel_bias,labeller = partial(label_both,sep=":\n"))+xlim(0,1)+ylim(0,1)
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
