---
title: "Simulating GWAS using snakemake"
author: "Nicholas Knoblauch"
date: 2017-09-18
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->



## Required software

1. `snakemake`
2. The following `R` packages:
  1) `RSSp` 
  2) `LDshrink` 
  3) `RcppEigenH5` (make sure you've loaded some kind of `HDF5` module)
  4) `SeqArray`


## Snakemake in 2 minutes

The idea behind snakemake is that instead of describing your workflow from beginning to end, you describe it in terms of inputs and outputs. "Rules", which map inputs to outputs are specified in files called `Snakefile` and are just plain python files that look something like this:

```{python,eval=F,echo=T}

all:
 input:
  "Y.RDS"

rule do_something:
  ''' rule do_something will be triggered when file Y.RDS
  does not exist, and it may in turn trigger the rule that
  generates file X.txt if it doesn't exist'''
  input:
    input_filename="X.txt"
  output:
    output_filename="Y.RDS"
  params:
    some_parameter=10
  script:
    "my_script.R"
    
rule do_something_else:
  ''' This rule uses the shell directly
  '''
  input:
    file_to_decompress="{wc}.tar.gz"
  output:
    decomp_file="{wc}.tar"
  shell:
  "gunzip {input.file_to_decompress}"
    
  
    
rule make_letter:
  ''' This rule takes no input!
  '''
  output:
    output_file="{wc}.txt"
  params:
    letter="{wc}"
  script:
    "make_letter.R"
    
    
```


In the same directory as `Snakefile` would be a script `my_script.R` which might look something like this 


```{r,eval=F,echo=T}

my_result <- length(scan(snakemake@input[["input_filename"]],what=character()))

my_result <- my_result*snakemake@params[["some_parameter"]]

saveRDS(my_result,snakemake@output[["output_filename"]])

```

and another script `make_letter.R` that might look like this:

```{r,eval=F,echo=T}


my_letter <- snakemake@params[["letter"]]

saveRDS(letters[1:which(letters==my_letter)],snakemake@output[["output_file"]])

```



When  the command `snakemake` is invoked in that directory, `snakemake` looks first at the rule `all`, sees that it takes as input `Y.RDS`, and will look in the directory for `Y.RDS`, if it finds it, then it will say `nothing to be done`.  If `Y.RDS` doesn't exist then `snakemake` will see that `Y.RDS` is made by the rule `do_something`, which needs `X.txt`.  If `X.txt` exists, then `snakemake` runs `my_script.R`, feeding in the parameters as needed. If `X.txt` doesn't exist then `snakemake` looks at the rule `make_letter`, and sees that if it substitutes `X` in `{wc}.txt` it will get `X.txt`, so it runs `make_letter`, then `do_something`, and in the end you have `Y.RDS`.


## Simulating GWAS (on RCC)

All of the rules for simulating data have a docstring that describe what they do.  Probably the most relevant ones will be `sim_pheno_w_Beta_RSSp` and `map_betahat_se`.

`sim_pheno_w_Beta_RSSp` takes as input: a SeqArray `gds` file (there should be one already on in `/project/xinhe/eQTL/1kg/gds/ALL_seq_hapmap_geno.gds`); an `RDS` file `betadf.RDS` and a vector of $PVE$ values (specified in the `Snakefile`).
`betadf.RDS` should contain a dataframe that has the following columns (it can contain more columns, but they'll be ignored):

1. `SNP` the rsid of the SNP
2. `beta` the true effect of the (scaled) dosage on the trait
3. `fgeneid` the name of the trait. 

`betadf.RDS` should have $p \times g$ rows, where $p$ is the number of SNPs in the simulation and $g$ is the number of traits in the simulation ($PVE$ should also be of length $g$)

The output of `sim_pheno_w_Beta_RSSp` is an HDF5 file with trait values stored in a matrix. (the matrix of trait values can be read with `ymat <- RcppEigenH5::read_2d_mat_h5("RSSp_sim_gwas_pheno/simulate_y.h5","trait","ymat")`. 

`map_betahat_se` takes a `gds` file (the same one used in `sim_pheno_w_Beta_RSSp`) and an HDF5 file (like the one generated by `sim_pheno_w_Beta_RSSp`, and returns a `.txt.gz` file with betahat,standard error for each snp and each trait.

## Running snakemake on RCC

The easiest way to run snakemake on RCC is to run everything in one interactive job.
The workflow looks like this:
0. Clone the `LDshrink` repo and change in to the directory  that contains the `Snakefile` (in this case, something ending in `LDshrink/inst/snakemake_files` )
1. Edit the json file `workflow_params.json` so that all the directories point to where you want them to point (there are some notes on this in the `Snakefile`).  Make sure that the directories are specified *relative* to the directory containing the `Snakefile` (i.e starting with `../` and not `/` ).
2. Request an interactive job using `sinteractive` or `srun`.
3. Set up modules that you can use `HDF5`,`R`, and `snakemake` (`python3`).
4. Change back to the `Snakefile` directory.
5. Use `snakemake -n` to see what files are slated to be created/which rules are set to be run.
6. Use `snakemake` (or `snakemake -j n` where `n` is the number of threads/cores/CPUs you want to use) to run the rules.






## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
