---
title: "UK Biobank PVE"
author: "Nicholas Knoblauch"
date: "2020-03-30"
output: workflowr::wflow_html
---

## Introduction

Below are pve estimates from 69 uk biobank traits for which I could easily obtain:
1) GWAS summary statistics (from BoltLMM) 
2) LD score regression heritability estimates 
3) Individual-level-data based heritability estimates (from mmhe, described [here](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1006711#sec004)) 

I ran `RSSp` for each of these 69 traits (using Xiang's new method for calculating pve).  First we see the summary based heritability estimates (y axis) versus individual level estimate (x-axis). 

**Update** 2020-03-13:  the link for ld score regression results has changed to "https://www.dropbox.com/s/ipeqyhrpdqav5uh/ukb31063_h2_all.02Oct2019.tsv.gz?dl=1"

univariate gwas links were obtained here: https://docs.google.com/spreadsheets/d/1EmwlCYYqkoVKqAS71nNDKoN18PyKwysYUppKTSMvKiM/edit#gid=450760074


```{r,echo=F,message=F,warning=F}
suppressMessages(library(plotly))
suppressMessages(library(tidyverse))
library(fs)
library(ldmap)
library(EigenH5)
library(forcats)
```



```{r, echo = FALSE, eval = FALSE}
all_pheno_url <- "https://www.dropbox.com/s/ipeqyhrpdqav5uh/ukb31063_h2_all.02Oct2019.tsv.gz?dl=1"
destf <- "~/Dropbox/PolygenicRSS/data/ukb31063_h2_all.02Oct2019.tsv.gz"
sumstatf   <- "~/Dropbox/PolygenicRSS/data/LDSC_UKB_SumStatManifest.tsv"
sumstat_dl_df <- readr::read_tsv(sumstatf)
download.file(all_pheno_url,destf)
```


```{r, echo = F, message = F, warning = F}
aukb <- "~/Dropbox/PolygenicRSS/code/snakemake_files/UKBB_mf.csv"
ukbf <- read_csv(aukb)
est_frds <- dir_ls("~/Dropbox/PolygenicRSS/output/pve/ukb/",
                   recurse = T,
                   type = "file",
                   glob = "*RDS")
est_df <- file_info(est_frds) %>%
    filter(size < fs::as_fs_bytes(1e6))
ukb_rssp_c <- map(est_df$path, safely(readRDS)) %>%
    map("result") %>%
    compact() %>%
    map(~ select(.x, -bias, -info)) %>%
    bind_rows() %>%
    rename(trait = trait_id) %>%
    filter(convergence == 0) %>%
    mutate(trait = str_remove(trait, "^ukb_")) %>%
    filter(nterms == 1) %>%
    rename(rssp = pve)


ukb_hdf <- readxl::read_xlsx("~/Downloads/journal.pgen.1006711.s002.xlsx") %>%
    select(trait = `Field ID`,
           pve_ind = h2,
           N_ind = N,
           trait_name = `Field Name`)

ukb_ldsc <- vroom::vroom(destf,delim = "\t")  %>%
    select(trait = phen_stem,
           desc_ldsc = description,
           trait_t = phenotype,
           ldsc_o = h2_observed,
           ldsc_l = h2_liability,
           N_ldsc = Neff,
           int_ldsc = intercept) %>%
    mutate(trait_t = factor(str_replace(trait_t, "^[0-9]+_(.+)", "\\1")))

                                                                    

ukb_all <- inner_join(ukb_ldsc, ukb_rssp_c, by = "trait") %>%
    inner_join(ukb_hdf, by = "trait") %>%
    distinct()

ukb_sp  <- gather(ukb_all,
                  key = "method",
                  value = "pve_estimate",
                  ldsc_o,
                  ldsc_l,
                  rssp)

ggplot(ukb_all, aes(x = pve_ind, y = pve)) + geom_point()
```

```{r}
library(RSSp)
test_res <- arrange(est_df,desc(size)) %>% slice(1) %>% pull(path) %>% readRDS()
saveRDS(test_res$data[[1]]$quh,"~/Desktop/test_quh.RDS")
saveRDS(test_res$data[[1]]$D,"~/Desktop/test_D.RDS")

test_dat <- test_res$data[[1]]
N <- test_res$info$median_N
trait <- test_res$trait_id
rssp_res <- RSSp_estimate(quh = test_dat$quh,
                          D = test_dat$D,
                          sample_size = N,
                          trait_id = trait,nterms = 2)
```


```{r}
est_f <- dir_ls("~/Dropbox/PolygenicRSS/output/pve/ukb/",recurse = T,type = "file",glob="*gz")
ukb_rssp_a <- map_df(est_f,~suppressMessages(read_delim(.x,delim="\t")))
```


```{r}
b_rssp <- rbind(select(ukb_rssp_c,one_of(intersect(colnames(ukb_rssp_c),colnames(ukb_rssp_a)))),
      select(ukb_rssp_a,one_of(intersect(colnames(ukb_rssp_c),colnames(ukb_rssp_a)))))


tb_rssp <- b_rssp %>% select(trait_id,pve,useConfounding) %>% 
  spread(key = useConfounding,value=pve)
ggplot(tb_rssp,aes(x=old,y=new))+geom_point()+geom_abline(slope=1,intercept=0)
```









```{r,echo=F,message=F,warning=F}
suppressMessages(library(plotly))
suppressMessages(library(tidyverse))
library(fs)
library(forcats)

est_frds <- dir_ls("~/Dropbox/PolygenicRSS/output/pve/ukb/",recurse = T,type = "file",glob="*RDS")

read_frds <- function(fn){
  
  td <- safely(readRDS)(fn) 
  if(is.null(td$result)){
    return(td$result)
  }
  td <- td$result
  class(td) <- c(class(td),class(tibble()))
  td <- td %>% select(-data)
  # td <- purrr::discard(td,)
  td$trait <- td$info$trait
  td$sample_size <- td$info$median_N
  td <- select(td,-info)
  return(td)
}
#trait_ctf <- dir_ls("~/Desktop/gardner_scratch/summary_statistics_fst/",glob="*sc.txt")
ukb_hdf <- readxl::read_xlsx("~/Downloads/journal.pgen.1006711.s002.xlsx") %>% 
  select(trait=`Field ID`,pve_ind=h2,N_ind=N,trait_name=`Field Name`)
ukb_ldsc <- read_tsv("~/Downloads/ukbb_all_h2univar_results.txt.gz") %>% select(trait=phenotype,pve_ldsc=h2_observed,N_ldsc=N,int_ldsc=intercept)

ukb_rssp <- map_dfr(est_frds,read_frds) 
ukb_rssp_c <- select(ukb_rssp,nterms,pve,trait) %>%
    mutate(nterms=paste0("pve_RSSP_",nterms)) %>%
    spread(nterms,pve) %>% 
  mutate(trait=as.character(trait))
  
all_traits <- inner_join(ukb_hdf, ukb_ldsc) %>%
  inner_join(ukb_rssp_c)

  #distinct(trait,=T) 
```


```{r}
nall_traits <- filter(all_traits,useConfounding=="2") %>% select(-useConfounding)
max_pve <- max(c(nall_traits$pve_ind,nall_traits$pve_ldsc,nall_traits$pve_rssp))
new_names <- c("pve_ldsc"="LDSC","pve_rssp"="RSSp")
new_df <- rename(nall_traits,sample_size=N_rssp) %>% 
  select(-starts_with("N_")) %>%
  gather(key = "method",value="pve_sumstat",-trait,-pve_ind,-int_ldsc,-sample_size) %>% 
  mutate(method=new_names[method]) 
new_df%>%  ggplot(aes(x=pve_ind,y=pve_sumstat,col=method))+geom_point()+geom_abline(slope=1,intercept=0)+scale_x_continuous("Individual-Level-Data PVE (mmhe)",limits=c(0,max_pve))+scale_y_continuous("Summary statistic PVE",limits=c(0,max_pve))+geom_smooth(method="lm",formula = y~x+0)+ggtitle("RSSp vs LDSC on UK biobank traits")
```


```{r}
ggplot(all_traits,aes(x=pve_ind,y=pve_ldsc))+geom_point()
```

```{r}
ggplot(all_traits,aes(x=pve_ind,y=pve_rssp))+geom_point()
```



```{r,echo=F,message=F,warning=F}
max_pve <- max(c(all_traits$pve_ind,all_traits$pve_ldsc,all_traits$pve_rssp))
new_names <- c("pve_ldsc"="LDSC","pve_rssp"="RSSp")
rename(all_traits,sample_size=N_rssp) %>% select(-starts_with("N_")) %>% gather(key = "method",value="pve_sumstat",-trait,-pve_ind,-int_ldsc,-sample_size) %>% mutate(method=new_names[method]) %>%  ggplot(aes(x=pve_ind,y=pve_sumstat,col=method))+geom_point()+geom_abline(slope=1,intercept=0)+scale_x_continuous("Individual-Level-Data PVE (mmhe)",limits=c(0,max_pve))+scale_y_continuous("Summary statistic PVE",limits=c(0,max_pve))+geom_smooth(method="lm",formula = y~x+0)+ggtitle("RSSp vs LDSC on UK biobank traits")
```

If we treat the individual-level estimates as "truth", `RSSp` has lower "error" on average.
```{r,echo=F,message=F,warning=F}
select(all_traits,-starts_with("N_")) %>%
  gather(key = "method",value="pve_sumstat",-trait,-pve_ind,-int_ldsc) %>%mutate(method=new_names[method]) %>% 
  ggplot(aes(x=method,y=abs(pve_sumstat-pve_ind),col=method))+geom_boxplot()+ggtitle("RSSp PVE estimates are closer to individual-level-data based estimates")
```

One of the first thoughts I had was that perhaps LDSC's intercept term is contributing to the underestimation of heritability. 
```{r,echo=F,message=F,warning=F}
rename(all_traits,sample_size=N_rssp) %>% select(-starts_with("N_")) %>%
  gather(key = "method",value="pve_sumstat",-trait,-pve_ind,-sample_size,-int_ldsc)%>%mutate(method=new_names[method]) %>% mutate(pve_level=cut_number(pve_ind,3)) %>% 
  ggplot(aes(x=int_ldsc,y=pve_sumstat-pve_ind,col=method))+geom_point()+xlab("LD score regression Intercept ")+geom_smooth(method="lm")+ggtitle("LDSC's Intercept term isn't helping LDSC")
```


It appears that `RSSp` is fairly consistently (but moderately) outperforming LDSC across sample size, and PVE level.


```{r,echo=F,message=F,warning=F}
rename(all_traits,sample_size=N_rssp) %>% select(-starts_with("N_")) %>%
  gather(key = "method",value="pve_sumstat",-trait,-pve_ind,-sample_size,-int_ldsc)%>%mutate(method=new_names[method]) %>% mutate(pve_level=cut_number(pve_ind,3)) %>% 
  ggplot(aes(x=log10(sample_size),y=abs(pve_sumstat-pve_ind)/pve_ind,col=method))+geom_point()+ggtitle("Relative Error as a function of sample size","Stratified by Individual level PVE")+geom_smooth(method="lm")+xlab("sample size (log10)")+facet_wrap(~pve_level,labeller = label_both)+ylab("Relative RMSE")
```





```{r}
library(EigenH5)
library(RSSp)

good_trait <- 1239
nterms <- 2
good_quhf <- glue::glue("~/Dropbox/scratch/polyg_scratch/quh/gwas_chr{1:22}AF0SNP0N0_ukb_EUR_T_EUR_gwas_T_0.h5")
trait_id <- EigenH5::read_vector_h5(good_quhf[1],"TraitInfo/trait")
trait_index <- which(trait_id==good_trait)

quh_df <- map_df(good_quhf,~data_frame(
  quh=c(read_matrix_h5(.x,"quh_df/quh",subset_cols=trait_index)),
  D=read_vector_h5(.x,"quh_df/D"))
  )
sc_df <- read_tsv(glue::glue("~/Dropbox/scratch/polyg_scratch/summary_statistics_fst/ukb_{good_trait}_summary_statistics_sc.txt"))
test_est <- RSSp_estimate(quh_df$quh,quh_df$D,sample_size=sc_df$median_N,nterms = 4)

```

<!-- ```{r} -->
<!-- bad_quhf <- fs::dir_ls("~/Desktop/gardner_scratch/quh/",glob = glue::glue("*AF0SNP0N0_ukb0{bad_trait}_EUR_T_EUR_gwas_T_0.h5")) -->
<!-- bad_quh_df <- map_df(bad_quhf,~EigenH5::read_df_h5(.x,"quh_df")) -->




<!-- ```{r} -->

<!-- library(RSSp) -->
<!-- good_quh_df <- filter(quh_df,!is.na(quh)) -->
<!-- rss_res <- RSSp:::RSSp_estimate(quh = good_quh_df$quh,D = good_quh_df$D,p_n = nrow(quh_df)/51453) -->

<!-- sum_quh_df <- group_by(quh_df,ntile(D^2,1000)) %>% summarise(md2=mean(D^2),mean_quh=mean(quh),mquhsq=mean(quh^2-D)) -->

<!-- ggplot(sum_quh_df,aes(x=md2,y=mquhsq))+geom_point() -->
<!-- ``` -->



<!-- ```{r} -->

<!-- trait_ls <- function(dir="~/Dropbox/PolygenicRSS/output/pve/"){ -->
<!--   est_f <- dir_ls(dir,recurse = T,type = "file") -->

<!-- } -->
<!-- ``` -->


<!-- ```{r} -->

<!-- quh_f <- function(consortia="angst",trait="anx",quh_d= "~/Dropbox/scratch/polyg_scratch/quh/"){ -->
<!--   fs::path_expand(glue("{quh_d}","gwas_chr{1:22}AF0SNP0N0_{consortia}0{trait}_EUR_T_EUR_gwas_T_0.h5")) -->
<!-- } -->

<!-- quhf <- quh_f() -->
<!-- read_quh_df <- function(quhf){ -->
<!--   map_df(quhf,read_df_h5,datapath="quh_df") -->

<!-- } -->

<!-- quh_df <- read_quh_df(quhf) -->
<!-- library(fs) -->
<!-- library(forcats) -->
<!-- library(ggplot2) -->
<!-- library(EigenH5) -->

<!-- rssr <- RSSp:::RSSp_estimate(mutate(quh_df,fgeneid=1),p_n=nrow(quh_df)/2000L) -->
<!-- slm <- lm(I(quh^2)~I(D^2)+offset(D)+0,data=quh_df) -->
<!-- quh_f <- dir_ls(quh_d,recurse = F,type = "file") -->
<!-- tquh <- read_df_h5(quh_f[1],"quh_df") -->
<!-- ggplot(tquh,aes(x=quh^2,y=D))+geom_point() -->
<!-- ``` -->

