---
  title: "UK Biobank PVE"
author: "Nicholas Knoblauch"
date: "2018-11-14"
output: workflowr::wflow_html
---

  ## Introduction

  Below are pve estimates from 69 uk biobank traits for which I could easily obtain:
  1) GWAS summary statistics (from BoltLMM)
2) LD score regression heritability estimates
3) Individual-level-data based heritability estimates (from mmhe, described [here](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1006711#sec004))

                                                                                   I ran `RSSp` for each of these 69 traits (using Xiang's new method for calculating pve).  First we see the summary based heritability estimates (y axis) versus individual level estimate (x-axis).




```{r,echo=F,message=F,warning=F}
suppressMessages(library(plotly))
suppressMessages(library(tidyverse))
library(fs)
library(forcats)

est_frds <- dir_ls("~/Dropbox/PolygenicRSS/output/pve/ukb/",recursive = T,type = "file",glob="*RDS")


ukb_rssp_c <- map(est_frds,~safely(readRDS)(.x)[["result"]]) %>% discard(~any(class(.x)=="rssp")) %>%discard(~is.null(names(.x))) %>%  map(~modify_at(.x,"info",function(x){
  return(list(x))
})) %>% bind_rows()
# ukb_rssp_df <- bind_rows(ukb_rssp_c)
# map(ukb_rssp_c,class)
ukb_rssp_c <- mutate(ukb_rssp_c,trait=as.integer(gsub("ukb_","",trait_id))) %>% select(-trait_id)
ukb_rssp_c <- mutate(ukb_rssp_c,nterms=map_int(bias,nrow)+1)

tukb <- ukb_rssp_c %>% filter(trait==trait_id[1])
#trait_ctf <- dir_ls("~/Desktop/gardner_scratch/summary_statistics_fst/",glob="*sc.txt")
ukb_hdf <- readxl::read_xlsx("~/Downloads/journal.pgen.1006711.s002.xlsx") %>% select(trait=`Field ID`,pve_ind=h2,N_ind=N)
ukb_ldsc <- read_tsv("~/Downloads/ukbb_all_h2univar_results.txt.gz") %>% select(trait=phenotype,pve_ldsc=h2_observed,N_ldsc=N,int_ldsc=intercept)
```

```{r}
all_traits <- inner_join(ukb_hdf, ukb_ldsc) %>% mutate(trait=as.integer(trait)) %>% 
  inner_join(ukb_rssp_c) %>% rename(pve_rssp=pve)
```

```{r}
p_traits <- select(all_traits,trait,starts_with("pve"),nterms)
ggplot()
max_pve <- max(c(p_traits$pve_ind,p_traits$pve_ldsc,p_traits$pve_rssp))
new_names <- c("pve_ldsc"="LDSC","pve_rssp"="RSSp")
new_df <- rename(p_traits) %>% 
  select(-starts_with("N_")) %>%
  gather(key = "method",value="pve_sumstat",-trait,-pve_ind,-nterms) %>% 
  mutate(method=new_names[method]) 
new_df%>%  ggplot(aes(x=pve_ind,y=pve_sumstat,col=method))+geom_point()+geom_abline(slope=1,intercept=0)+scale_x_continuous("Individual-Level-Data PVE (mmhe)",limits=c(0,max_pve))+scale_y_continuous("Summary statistic PVE",limits=c(0,max_pve))+geom_smooth(method="lm",formula = y~x+0)+ggtitle("RSSp vs LDSC on UK biobank traits")+facet_wrap(~nterms)
```




```{r}
filter(new_df,method!="LDSC") %>% ggplot(aes(x=pve_ind,y=pve_sumstat,col=factor(nterms)))+geom_point()+geom_abline(slope=1,intercept=0)+scale_x_continuous("Individual-Level-Data PVE (mmhe)",limits=c(0,max_pve))+scale_y_continuous("Summary statistic PVE",limits=c(0,max_pve))+geom_smooth(method="lm",formula = y~x+0)+ggtitle("RSSp vs LDSC on UK biobank traits")
```


```{r}
ukb_rssp_c <- inner_join(ukb_rssp_c,mutate(ukb_hdf,trait=as.integer(trait)))

bias_rssp <- unnest(ukb_rssp_c,bias)


```



